---
title: "Introduction to geno2ped"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geno2ped}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`geno2ped` is an R package for rapid pedigree construction from SNP genotype 
data. It infers parent-offspring relationships using genomic similarity and 
Mendelian error validation, and supports PLINK, VCF, and CSV input formats.

## Installation

```{r install, eval = FALSE}
devtools::install_github("ampofoissabel-quangen/geno2ped")
```

## Loading the Package

```{r load}
library(geno2ped)
```

## Reading Genotypes

### From a CSV file
The simplest input format is a CSV where the first column contains sample IDs
and the remaining columns contain SNP genotypes (0/1/2):

```{r read-csv, eval = FALSE}
geno <- read_genotypes(
  csv             = "path/to/genotypes.csv",
  sample_metadata = "path/to/metadata.csv"
)
```

### From PLINK files
```{r read-plink, eval = FALSE}
geno <- read_genotypes(
  bed_prefix      = "path/to/mydata",
  sample_metadata = "path/to/metadata.csv"
)
```

### From VCF files
```{r read-vcf, eval = FALSE}
geno <- read_genotypes(
  vcf             = "path/to/mydata.vcf.gz",
  sample_metadata = "path/to/metadata.csv"
)
```

### The genotype object
`read_genotypes()` returns a list with four elements:

| Element  | Description                              |
|----------|------------------------------------------|
| `$G`     | Genotype matrix (samples x SNPs, 0/1/2) |
| `$ids`   | Sample ID vector                         |
| `$map`   | SNP map (chr, pos, snp)                  |
| `$meta`  | Sample metadata (ID, Sex, BirthYear)     |

### Sample metadata format
Your metadata CSV should look like this:

| ID        | Sex | BirthYear |
|-----------|-----|-----------|
| Sample001 | M   | 1985      |
| Sample002 | F   | 1990      |
| Sample003 | M   | 2005      |

## Building a Pedigree

### Using a preset
The easiest way is to use a built-in preset:

```{r build, eval = FALSE}
res <- build_pedigree(
  geno,
  preset  = "high_precision",
  verbose = TRUE
)
```

Three presets are available:

| Preset           | Use case                              |
|------------------|---------------------------------------|
| `none`           | Manual threshold control              |
| `high_precision` | Minimise false positives              |
| `high_coverage`  | Maximise assignments                  |

### Custom thresholds
```{r custom, eval = FALSE}
res <- build_pedigree(
  geno,
  s_threshold    = 0.75,
  me_max         = 0.003,
  oh_max         = 20,
  use_sex        = TRUE,
  use_age        = TRUE,
  min_parent_gap = 2,
  top_k          = 3
)
```

### The result object
`build_pedigree()` returns a list with:

| Element       | Description                              |
|---------------|------------------------------------------|
| `$pedigree`   | Final pedigree (ID, Sire, Dam, status)   |
| `$assignments`| All candidate parent scores              |
| `$trios`      | Best trio per offspring                  |
| `$trios_all`  | All evaluated trios                      |
| `$summary`    | Assignment statistics                    |
| `$settings`   | Parameters used                          |

### Viewing results
```{r results, eval = FALSE}
# View the pedigree
res$pedigree

# View summary statistics
res$summary
```

## Visualising Results

```{r plots, eval = FALSE}
# Similarity score distribution
plot_kinship(res)

# Mendelian error rate distribution
plot_me(res)

# Assignment outcome counts
plot_assignment_status(res)
```

## Saving Results

```{r save, eval = FALSE}
write_outputs(
  res,
  outdir = "geno2ped_results",
  prefix = "my_run"
)
```

This creates the following files in `geno2ped_results/`:

- `my_run_pedigree.csv` — final pedigree
- `my_run_assignments.csv` — candidate scores
- `my_run_trios.csv` — best trios
- `my_run_summary.txt` — run statistics
- `my_run_settings.txt` — parameters used

## Session Info

```{r session}
sessionInfo()
```
